---
title: "NYPD Shooting Incident Data (Historic)"
author: "Sam Asemi"
date: "2025-01-20"
output:
  html_document: default
  pdf_document: default
---



## Project Overview

This analysis examines shooting incidents in New York City based on historic data from the NYPD. We will explore trends in incidents over the years and by region, along with a demographic breakdown of victims and shooters.

## Data Source

The data used in this analysis is retrieved from the NYC Open Data portal:
- **Dataset URL:** [NYPD Shooting Incident Data](https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD)

```{r}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
```

### Loading necessary libraries
```{r message=FALSE}
library(readr)  
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
library(forecast)
```

```{r }
# Load the data from the URL
nypd_data <- read_csv(url, show_col_types = FALSE)
```

### Data Cleaning and Preparation


#### Removing unnecessary columns and rename for clarity

```{r message=FALSE}
duplicates <- nypd_data %>%
  count(INCIDENT_KEY) %>%
  filter(n > 1) %>%
  nrow()
print(paste("Total duplicate records:", duplicates))

# Number of rows before removing duplicates
nrow(nypd_data) 

# Removing duplicate rows
nypd_data <- nypd_data %>%
  distinct(INCIDENT_KEY, .keep_all = TRUE)

# Number of rows after cleaning duplicates
nrow(nypd_data)  

```

```{r message=FALSE}
# Removing unnecessary columns and rename for clarity
nypd_data <- nypd_data %>% 
  select(-INCIDENT_KEY, 
         -PRECINCT, 
         -JURISDICTION_CODE,
         -LOC_OF_OCCUR_DESC,
         -LOCATION_DESC,
         -X_COORD_CD, 
         -Y_COORD_CD, 
         -Latitude, 
         -Longitude, 
         -Lon_Lat,
         -OCCUR_TIME) %>%
  rename(
    Incident_date = OCCUR_DATE,
    Region = BORO,
    Location = LOC_CLASSFCTN_DESC,
    Murder = STATISTICAL_MURDER_FLAG,
    Shooter_age_group = PERP_AGE_GROUP,
    Shooter_sex = PERP_SEX,
    Shooter_race = PERP_RACE,
    Victim_age_group = VIC_AGE_GROUP,
    Victim_sex = VIC_SEX,
    Victim_race = VIC_RACE
  ) %>%
  mutate(Incident_date = mdy(Incident_date))

# Preview the summarized data
head(nypd_data)
```
## Analysis
### Shooting Incidents Trend Over Years

```{r warning=FALSE, message=FALSE}

# Add a column for the year and summarize incidents by year and region
nypd_by_year <- nypd_data %>%
  mutate(Year = year(Incident_date)) %>%
  group_by(Year, Region) %>%
  summarise(Count = n())

# Total Crimes Per Year
crime_per_year <- nypd_data %>%
  group_by(Year = year(Incident_date)) %>%
  summarise(Crime_Number = n(), .groups = "drop")

# Preview the summarized data
head(nypd_by_year)
```

### 1- Line Chart of Number Incidents by Year

#### The following chart visualizes the trend and number of shooting incidents over the years in NY state.

```{r}

ggplot(crime_per_year, aes(x = Year , y = Crime_Number)) +
  geom_line(color = "#1f77b4", linewidth = 1) +
  ggtitle("Total Crimes Per Year") +
  xlab("Year") + ylab("Number of Crimes") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = 'bold'))
```

### 2- Heatmap of Incidents by Year and Region

#### The following heatmap visualizes shooting incidents across different regions over the years.

```{r }
nypd_by_year %>%
  ggplot(aes(x = as.factor(Year), y = Region, fill = Count)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "lightblue", high = "darkblue") +
    labs(
      title = "Incidents by Year and Region",
      x = "Year",
      y = "Region",
      fill = "Incident Count"
    ) +
    theme_minimal()+
    theme(plot.title = element_text(hjust = 0.5, face = 'bold'), 
          axis.text.x = element_text(angle = 45, hjust =1)
          )

```


### 3- Line Chart of Incidents Over Years by Region

#### The line chart below shows the trends in shooting incidents across different regions over time.

```{r warning=FALSE}
nypd_by_year %>%
  ggplot(aes(x = Year, y = Count, color = Region, group = Region)) +
    geom_line(size = 1) +
    geom_point(size = 3) +
    labs(
      title = "Incidents Over Years by Region",
      x = "Year",
      y = "Incident Count",
      color = "Region"
    ) +
    theme_minimal()+
    theme(plot.title = element_text(hjust = 0.5, face = 'bold'))

```


### 4- Stacked Bar Chart of Incidents by Shooter Age Group and Region from 2006 to 2023

#### This visual illustrates the distribution of percentage of different shooter age groups in all the boroughs in NY state over the years.
```{r warning=FALSE, message=FALSE}

# Add a column for the year and summarize incidents by shooter age group and region
nypd_by_age <- nypd_data %>%
  mutate(Year = year(Incident_date)) %>%
  group_by(Shooter_age_group, Region) %>%
  summarise(Count = n())
# Removing null values
nypd_by_age <- nypd_by_age %>% drop_na()
nypd_by_age <- nypd_by_age[-c(1,2,3,4,5,6,7,11,13,34,35,36,37,38), ]
nypd_by_age_percent <- nypd_by_age %>%
  group_by(Region) %>%
  mutate(Percent = Count / sum(Count) * 100) %>%
  ungroup()

# Stacked bar chart
ggplot(nypd_by_age_percent, aes(x = Region, y = Percent, fill = Shooter_age_group)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.8) +  # Add transparency to bars
  labs(
    title = "Incidents by Shooter Age Group and Region 2006 - 2023",
    x = "Region",
    y = "Percentage of Incidents",
    fill = "Shooter Age Group",
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = 'bold'),  
    legend.position = "right",  
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),  
    axis.text.x = element_text(angle = 45, hjust = 1),  
    plot.caption = element_text(hjust = 1, size = 9, color = "gray50")  
  ) +
  scale_fill_brewer(palette = "Set2") +  
  geom_text(
    aes(label = ifelse(Percent > 5, paste0(round(Percent, 1), "%"), "")), 
    position = position_stack(vjust = 0.5), size = 3.5, color = "black", fontface = "bold"
  ) +
  geom_segment(
    aes(x = Region, xend = Region, y = 0, yend = 100),   
    color = "gray50", linetype = "dotted", size = 0.3
  )
```

### Fitting a linear model to predict the number of incidents based on the available data in the future years.
```{r}
# Fit a linear model to predict total crimes per year
crime_model <- lm(Crime_Number ~ Year, data = crime_per_year)

# Create a data frame with future years
future_years <- data.frame(Year = c(2025, 2030, 2035))

# Predict number of crimes for future years
predictions <- predict(crime_model, newdata = future_years)

# Combine results
future_years$Predicted_Crimes <- predictions

# View predictions
print(future_years)
```
### 5- The Linear Model Visual
#### This visual shows the prediction of the future of number of incidents that might happen in NY state.

```{r message=FALSE}

ggplot(crime_per_year, aes(x = Year, y = Crime_Number)) +
  geom_point(color = "#1f77b4", size = 2) +
  geom_smooth(method = "lm", color = "blue", linetype = "dashed") +
  geom_point(data = future_years, aes(x = Year, y = Predicted_Crimes), color = "red", size = 3) +
  ggtitle("Crime Predictions") +
  xlab("Year") +
  ylab("Number of Incidents") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, face = 'bold')) 
  
```

# Conclusion

* **Overall Trend in NY State**:

The number of incidents in New York State exhibits a general decreasing trend over time. However, a sharp increase was observed during the COVID-19 pandemic, followed by a subsequent decline.

* **Borough-Level Analysis**:

The data reveals that **Brooklyn** and **Bronx** consistently report the highest number of incidents, driving the overall trend observed in the visualizations.

* **Shooter Age Distribution**:

The stacked bar chart highlights that the majority of shooters fall within the **25-44** age group, followed by the **18-24** age group. This distribution, however, varies across boroughs, indicating regional differences in shooter demographics.

* **Future Projections**:

Based on the predictive model, the number of incidents in New York State is expected to continue declining, suggesting a positive trend in the coming years.


## Bias

* The New York Police Department (NYPD) is responsible for recording crime data, and their methods of reporting may introduce bias.

* The dataset only includes crimes reported to law enforcement.

* Some crimes may go unreported in some neighborhoods due to distrust in law enforcement, while others may be overreported in areas with higher police presence.

* Changes in law enforcement policies over time may affect reporting. Temporal Bias (Seasonality and Social Events)

* Crime rates might be influenced by seasonal changes (e.g., summer months may see more violent crimes due to increased outdoor activity).

* If police focus more on certain crimes (e.g., violent crimes), data may over represent specific categories.

